---
title: Stat 405 Mortality
author: Meera Borle, Isis Burgos, Naomi Consiglio, Carson Foster, Aidan Gerber
date: November 23, 2022
output:
    ioslides_presentation:
        widescreen: true
---
```{r setup, include=FALSE}
source("setup.r")
```

# Data

## Data Introduction
- Gathered from the National Bureau of Economic Research, who creates and distributes a dataset of US mortality for every year since 1959
- Each row represents a single death, while each column represents a different demographic characteristic of the deceased.
- Used the 2019 edition of the dataset since we did not want to focus on COVID-19
- Important information includes education, sex, age classification, day of month, place of death, weekday, manner of death, cause of death, and different - risk factors for each deceased individual


## Secondary Dataset
- Using Behavioral Risk Factor Surveillance System Survey
- Includes different free text survey questions from across the United States and territories with responses broken out by subgroup
    -- Includes questions on demographic characteristics, plus queries on current health behaviors, such as tobacco use and seatbelt use
- Combined the secondary dataset by matching up subgroups between the death dataset and the risk factor dataset 
    -- Tried to use aggregate statistics to analyze how risk factors can be matched with causes of death.


# Questions

## Given someone is dead, how did they die?

## Hypothesis + More Questions

# Exploration & Methods

---

```{r, weekday_deaths, fig.cap="caption", echo=FALSE}
source("plots/weekday_deaths.r", local = TRUE, print.eval = TRUE)
deaths_by_day_plot()
```

---

```{r weekday_deaths_scaled, fig.cap="caption", echo=FALSE}
deaths_by_day_plot_scaled()
```

---

```{r, months_deaths, fig.cap="caption", echo=FALSE}
source("plots/month_count.r", local = TRUE, print.eval = TRUE)
```

---

```{r, educ_cause, echo=FALSE, fig.width=10, fig.height = 6}
source("plots/educ_cause.r", local = TRUE, print.eval = TRUE)
plot_educ_cause()
```

## Table {.smaller}

```{r, educ_cause_table, echo=FALSE}
educ_cause_table()
```

---

```{r, killer, fig.cap="caption", echo=FALSE}
source("plots/killer.r", local = TRUE, print.eval = TRUE)

draw_killer(person_colors)
```

# Analysis

---

```{r, cancer, fig.cap="caption", echo=FALSE}
source("plots/cancer.r", local = TRUE, print.eval = TRUE)
```


