---
title: Stat 405 Mortality
author: Meera Borle, Isis Burgos, Naomi Consiglio, Carson Foster, Aidan Gerber
date: November 23, 2022
output:
    ioslides_presentation:
        widescreen: true
        css: style.css
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, out.width = "100%", fig.width = 9, fig.height = 5)
source("setup.r")

plotter <- function(p) {
    use_plotly(p)
}
```

# Data

## Data Introduction
- Gathered from the National Bureau of Economic Research, who creates and distributes a dataset of US mortality for every year since 1959
- Each row represents a single death, while each column represents a different demographic characteristic of the deceased.
- Used the 2019 edition of the dataset since we did not want to focus on COVID-19
- Important information includes education, sex, age classification, day of month, place of death, weekday, manner of death, cause of death, and different ailments for each deceased individual

## Secondary Dataset
- Using Behavioral Risk Factor Surveillance System Survey
- Includes different free text survey questions from across the United States and territories with responses broken out by subgroup
    - Includes questions on demographic characteristics, plus queries on current health behaviors, such as tobacco use and seatbelt use
- Combined the secondary dataset by matching up subgroups between the death dataset and the risk factor dataset 
    - Tried to use aggregate statistics to analyze how risk factors can be matched with causes of death.
## Given someone is dead, how did they die?
- Since we know that all of the people accounted for in this data are dead, we want to see if we can examine factors such as an individual's:
  - Age
  - Gender
  - Place of death
  - Educational Level
  - Health Conditions
  - Race

## What explains trends or irregularities in mortality when looking at different data factors?
- What specific factors mentioned previously are correlated with higher death rates.

# Exploration

---

```{r, weekday_deaths}
source("plots/weekday_deaths.r", local = TRUE, print.eval = TRUE)
plotter(deaths_by_day_plot())
```

---

```{r weekday_deaths_scaled}
plotter(deaths_by_day_plot_scaled())
```

---

```{r, months_deaths}
source("plots/month_count.r", local = TRUE, print.eval = TRUE)

plotter(month_count_plot(df_month_counts))
```

---

```{r, month_manner}
source("plots/month_manner.r", local = TRUE, print.eval = TRUE)
plotter(month_manner_plot())
```

--- 

```{r plot_age_gender}
source("plots/age_gender.r", local = TRUE, print.eval = TRUE)
plotter(deaths_by_age_gender_plot())
```

---

```{r, educ_cause, fig.height = 5.5}
source("plots/educ_cause.r", local = TRUE, print.eval = TRUE)
plotter(educ_cause_plot())
```

## Table {.smaller}

```{r, educ_cause_table}
educ_cause_table()
```

---

```{r killer}
source("plots/killer.r", local = TRUE, print.eval = TRUE)
# source("shiny.r", local = TRUE, print.eval = TRUE)

draw_killer(person_colors)
```

# Analysis

---

```{r cancer}
source("plots/cancer.r", local = TRUE, print.eval = TRUE)
plotter(cancer_free_text_plot())
```

---

```{r analysis, fig.cap="Decision Tree", message=FALSE}
# source("analysis.r", local = TRUE)
# accuracy <- run_decision_tree()
accuracy <- 0.8
```
`r 100 * accuracy`%

---

```{r analysis_random_forest, fig.cap="Random Forest"}
## accuracy <- run_random_forest()
```
TODO: fix random forest.
`r 100 * accuracy`%
